---
title: Announcing pins for Python
authors:
- Michael Chow
date: '2022-06-02'
slug: pins-for-python
blogcategories:
  - Open Source
tags:
  - pins
  - python
description: We are so happy to announce the release of pins for Python. You can use pins to publish, version, and share objects such as datasets and models.
alttext: "Two hex stickers: the pins hex sticker, orange with a pin, the Python logo, two intertwining snakes. The background is a fuzzy collage of other hex stickers."
events: blog
---


```{r}
#| echo: false
reticulate::use_virtualenv("~/.virtualenvs/pins-python")
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

We're excited to announce the release of [pins](https://rstudio.github.io/pins-python/) for Python!

pins removes the hassle of managing data across projects, colleagues, and teams by
providing a central place for people to store, version and retrieve data.
If you've ever chased a CSV through a series of email exchanges, or had to decide between
`data-final.csv` and `data-final-final.csv`, then pins is for you.

pins stores data on a **board**, which can be a local folder, or on RStudio Connect or a cloud provider like Amazon S3.
Each individual object (such as a dataframe, model, or another pickle-able Python object), together with some metadata, is called a pin.

The Python pins library works with its [R counterpart](https://pins.rstudio.com/),
so that teams working across R and Python have a unified strategy for sharing data.
This work emerged as part of RStudio's investment in Python open source, in order to
support bilingual data science teams.

<h2>Getting Started</h2>

The first step to using pins is installing it from PyPI.

```shell
python -m pip install pins
```

In the examples below, I'll walk through the basics of pins using a temporary directory
for a board, with `board_temp()`. This gets deleted after you close Python, so it is
not ideal for collaboration! You can use other boards, like `board_rsconnect()`, 
`board_folder()`, and `board_s3()`, in more realistic settings.


```{python}
import pins
from pins.data import mtcars

board = pins.board_temp()
```

You can "pin" (save) data to a board with the `.pin_write()` method. It requires three
arguments: an object, a name, and a pin type:

```{python}
board.pin_write(mtcars.head(), "mtcars", type="csv")
```

Above, we saved the data as a CSV, but depending on
what you’re saving and who else you want to read it, you might use the
`type` argument to instead save it as a `feather`, `parquet`, or `joblib` file.

You can later retrieve the pinned data with `.pin_read()`:

```{python}
board.pin_read("mtcars")
```

You can search for data using `.pin_search()` and `.pin_list()`.

```{python}
# prints out a list of all pins
# board.pin_list()

# searches for pins containing "cars"
board.pin_search("cars")
```

Two more pieces of important functionality exist:

* `.pin_write()` won't delete existing data, but versions your data.
* `.pin_read()` caches your data, so subsequent reads are much faster.

See [getting started](https://rstudio.github.io/pins-python/getting_started.html) in the
pins documentation for more information.

<h2>Interoperability with R pins</h2>

Pins stored with Python can be read with R, and vice-versa.

For example, here is R code that reads the `mtcars` pin we wrote to the board above.
Note that `TEMP_PATH` refers to the temporary directory we created in this blog post for our Python board.

```{r, echo=FALSE, results="hide"}
TEMP_PATH <- reticulate::py$board$board
```


```{r}
library(pins)

board <- board_folder(TEMP_PATH)
board %>% pin_read("mtcars")
```

This is especially useful when colleagues prefer one language over the other. For real collaborative work like this, you would use a board like `board_rsconnect()` or `board_s3()`.

<h2>Going further</h2>

The real power of pins comes when you share a board with multiple people.
To get started, you can use `board_folder()` with a directory on a shared
drive or in DropBox, or if you use
[RStudio Connect](https://www.rstudio.com/products/connect/) you can use
`board_rsconnect()`:

```python
board = pins.board_rsconnect()
board.pin_write(tidy_sales_data, "michael/sales-summary", type="csv")
```

Then, someone else (or an automated report) can read and use your
pin:

```python
board = pins.board_rsconnect()
board.pin_read("michael/sales-summary")
```

The pins package also includes boards that allow you to share data on
services like Amazon’s S3 (`board_s3()`), with plans to support other backends such as Google Cloud Storage and Azure's blob storage.


<h2>Get in touch</h2>

We are so happy about releasing pins for Python, and we want
to make sure it supports your workflow. Join our discussion on
RStudio Community to let us know what you're working on,
and how pins could help!
